\subsection{Documentation}
% You can place this under an appropriate \section{} or \subsection{} tag
% \section{Documentation Structure and Pedagogy}

\subsubsection{NeTEx Training Guide}

The documentation component is a progressive learning path for public transport professionals to NeTEx. It is designed as a self-contained training curriculum and organized into nine sequential modules that build incrementally upon one another. Each module pairs theoretical explanations with annotated XML examples derived from real operational data, specifically the STCP Line 200 (Bolhão--Castelo do Queijo) in Porto, ensuring that every concept is grounded in a real world context.

The pedagogical strategy was inspired by \cite{etalab_netex_formation} and was anchored in a consistent comparative approach: for every NeTEx construct introduced, the corresponding GTFS equivalent is presented side-by-side, explicitly highlighting what NeTEx adds in terms of expressiveness, interoperability, and regulatory compliance. This allows practitioners already familiar with GTFS (from previous experience or from the gtfs introduction documentation in the same repository) to leverage their knowledge while systematically acquiring the more rigorous data modeling capabilities that NeTEx demands.

The modules are organized as follows:

\begin{enumerate}
    \item \textbf{Prerequisites and Context:} Establishes the foundational context for the entire curriculum. It defines NeTEx and its relationship to the Transmodel conceptual framework (CEN EN 12896), situates the standard within the broader European ecosystem (Transmodel $\rightarrow$ NeTEx $\rightarrow$ SIRI $\rightarrow$ NAP), and points the regulatory obligations imposed by EU Delegated Regulation 2024/490 (MMTIS). A comparison between GTFS and NeTEx is also provided across multiple dimensions, format, conceptual model, coverage, identifiers, versioning, and extensibility, giving readers a overview before engaging with the technical content.

    \item \textbf{Core Concepts:} Introduces the structural foundations of NeTEx, starting from the VersionFrame architecture that organizes data into typed containers ({ResourceFrame, ServiceFrame, SiteFrame, TimetableFrame, FareFrame), and the hierarchical identifier and versioning system that ensures global uniqueness and temporal traceability. The module demonstrates how a NeTEx file is structured from the PublicationDelivery envelope down to individual entity references.

    \item \textbf{Transport Network:} Covers the modeling of transport network topology, decomposing it into three distinct layers: the commercial layer (Line, Direction), the topological layer (Route, RoutePoint, ServiceJourneyPattern), and the physical layer (StopPlace, Quay). This module explains the separation between logical stop points (ScheduledStopPoint) and physical infrastructure, a distinction absent in GTFS, and provides a complete mapping of routes.txt and stops.txt into their NeTEx equivalents.

    \item \textbf{Timetables and Calendars:} Addresses the modeling of vehicle journeys and service calendars. It introduces the ServiceJourney entity with its associated TimetabledPassingTime records, and the calendar system comprising DayType, OperatingDay, and DayTypeAssignment. The module demonstrates how trips.txt, stop\_times.txt, and calendar\_dates.txt from GTFS translate into the NeTEx TimetableFrame and ServiceCalendarFrame, emphasizing the reusability of journey patterns and the expressiveness of the model for representing complex service variations.

    \item \textbf{Multimodality and Physical Infrastructure:} Explores the NeTEx approach to modeling physical transport infrastructure through the SiteFrame. It details the StopPlace $\rightarrow$ Quay $\rightarrow$ BoardingPosition hierarchy, the PassengerStopAssignment mechanism that bridges logical and physical layers, and the use of PathLink elements for modeling pedestrian transfer paths between modes. Using the Bolhão intermodal node in Porto (bus and metro) as a case study, the module demonstrates how NeTEx enables precise multimodal interchange modeling that is structurally impossible in GTFS.

    \item \textbf{Accessibility:} Presents the NeTEx multi-dimensional accessibility model, which evaluates vehicle and infrastructure constraints independently through the AccessibilityAssessment and AccessibilityLimitation elements. Unlike the binary wheelchair\_accessible / wheelchair\_boarding fields in GTFS, NeTEx supports granular assessments across six dimensions (wheelchair access, step-free access, escalator-free access, lift-free access, audible signals, and visual signs), each with four possible states (yes, no, partial, unknown). The module also introduces the European EPIAP profile (NeTEx Part 6) and its regulatory implications for Portuguese operators.

    \item \textbf{Fare Structures (Introduction):} Provides an introductory overview of the NeTEx fare modeling system (Part 3). It covers the FareFrame structure, TariffZone definitions, PreassignedFareProduct entities for fixed-price tickets, and SalesOfferPackage elements for distribution channels. The Porto Andante ticketing system is used as the primary example, illustrating zone-based fare calculation. The module explicitly delineates its introductory scope, noting that advanced fare modeling (monthly passes, concessions, multi-operator interoperability) falls within the full NeTEx Part 3 specification.

    \item \textbf{Portuguese National NeTEx Profile:} Documents the Portuguese National Profile (PT-EPIP), the country-specific subset of NeTEx mandated by the IMT (Instituto da Mobilidade e dos Transportes) for data submission to the National Access Point (NAP Portugal). The module catalogs mandatory, recommended, and optional fields for each frame type, specifies the IFOPT-based identifier conventions, describes the expected file packaging structure for NAP submission, and provides a comparative analysis with the French national profile. This module addresses a gap in existing NeTEx educational resources, as no equivalent Portuguese-specific documentation was previously available.

    \item \textbf{Practical Tooling:} Equips readers with the practical tools and workflows necessary for production deployment. It covers NeTEx validation using the Entur netex-validator, XML editing environments (VS Code with the Red Hat XML extension, Oxygen XML Editor), GTFS-to-NeTEx conversion tools and their known limitations, XSD schema navigation techniques, and a comprehensive pre-submission checklist for the NAP Portugal. The module bridges the gap between understanding the standard and successfully implementing it in an operational context.
\end{enumerate}


Each module includes worked XML examples with extensive inline commentary, practical exercises with annotated solutions, and a formal references section citing all standards, regulations, and data sources used. By combining rigorous standards-based content with an accessible, GTFS-comparative pedagogy and real Portuguese operational data, the documentation enables transport operators and technical practitioners to confidently navigate the transition from GTFS to NeTEx in compliance with the Portuguese National Profile and European interoperability requirements.

\subsubsection{GTFS Training Guide}

The GTFS documentation component follows a parallel progressive structure, serving as a self-contained introduction to the General Transit Feed Specification for public transport professionals and engineering students with no prior GTFS experience. Organized into nine sequential modules (0--8), it constitutes the first open educational resource on GTFS authored in Portuguese, and is released under Creative Commons CC BY~4.0. As with the NeTEx documentation, every concept is anchored to real operational data from STCP Line~200 (Bolh\~ao--Castelo do Queijo) in Porto, ensuring that each file format, field, and modeling decision is illustrated in a recognizable and verifiable context.

The pedagogical approach is problem-driven: each module opens by establishing \emph{why} the constructs it introduces are necessary before presenting their technical specification. Optional NeTEx bridge sections appear throughout, mapping every GTFS primitive to its NeTEx counterpart and anticipating the transition that Portuguese operators must eventually undertake under EU Delegated Regulation~2024/490 (MMTIS). This bidirectional design makes the two documentation components mutually reinforcing rather than self-contained silos.

The modules are organized as follows:

\begin{enumerate}
    \item \textbf{Prerequisites and Context:} Establishes the origin and institutional history of GTFS (TriMet and Google, 2005; currently maintained by MobilityData), its global adoption across more than 10{,}000 agencies in over 100 countries, and its role within the European regulatory landscape, specifically the MMTIS obligations and the Portuguese National Access Point (NAP Portugal) managed by IMT. The module situates GTFS relative to NeTEx, characterising it as the practical baseline for feed publication rather than a permanent endpoint, and guides readers through the tool setup required for subsequent exercises.

    \item \textbf{Core Concepts:} Presents the relational data model that underpins the entire specification. GTFS is framed as a relational database exported to CSV, and the hierarchical chain from agency through routes and trips down to individual stop times is established through entity-relationship diagrams. The module covers the six mandatory files, data type conventions, the UTF-8 encoding requirement, and the non-intuitive time format that permits values beyond \texttt{24:00:00} to represent post-midnight service without a date change.

    \item \textbf{Agency, Stops, and Routes:} Provides a detailed examination of the three foundational tables. For \texttt{agency.txt}, the module covers operator metadata and timezone handling. For \texttt{stops.txt}, it explains the five-level location hierarchy (\texttt{location\_type} 0--4), parent-station inheritance, wheelchair boarding flags, and zone assignments used by the fare model. For \texttt{routes.txt}, it details route type codes, colour conventions, and naming best practices, using Line~200 as the running example throughout.

    \item \textbf{Trips, Timetables, and Geometry:} Addresses the service-instance layer of the model. It covers \texttt{trips.txt} fields including \texttt{direction\_id}, \texttt{block\_id} for vehicle blocking, and accessibility flags; \texttt{stop\_times.txt} with \texttt{timepoint} exactness coding and pickup/drop-off type enumeration; \texttt{shapes.txt} for geographic route traces; and \texttt{frequencies.txt} as an alternative representation for headway-based service. Particular attention is given to the midnight-crossing convention and to the practical importance of shapes for consumer map rendering despite their optional status in the specification.

    \item \textbf{Calendars and Exceptions:} Explains how GTFS decouples service patterns from their temporal validity through the \texttt{service\_id} abstraction. The module presents both modelling strategies---rule-based (\texttt{calendar.txt}) and exception-based (\texttt{calendar\_dates.txt})---and examines the trade-offs between human readability and programmatic generation. The actual STCP approach, which uses \texttt{calendar\_dates.txt} exclusively with \texttt{exception\_type=1}, is documented as a concrete example, alongside a reference table of Portuguese public holidays and guidance on feed validity windows.

    \item \textbf{Accessibility and Transfers:} Covers the two primary mechanisms for modelling physical usability and intermodal connectivity. Accessibility fields in \texttt{stops.txt} and \texttt{trips.txt} are presented alongside their inheritance rules. The \texttt{transfers.txt} file is explained through the Bolh\~ao intermodal node (Metro Linha~D and STCP), with asymmetric transfer times reflecting the physical geometry of the station. The module also introduces \texttt{pathways.txt} and \texttt{levels.txt} for detailed pedestrian infrastructure modelling, and situates these fields within the accessibility obligations imposed by EU Regulation~2024/490 and the Portuguese Decree-Lei~163/2006.

    \item \textbf{Data Lifecycle:} Traces a GTFS feed from source systems through production and publication. It surveys data inputs (planning software such as HASTUS and Optibus, AVL systems, and municipal geospatial data), the extract-transform-load pipeline structure, the canonical MobilityData GTFS Validator with its three severity levels (ERROR, WARNING, INFO), and feed publication to NAP Portugal and the Mobility Database. Practical guidance on feed versioning, update frequency, expiration monitoring, and the five quality dimensions (completeness, accuracy, timeliness, consistency, and availability) is provided.

    \item \textbf{Fares and Extensions:} Introduces the two generations of GTFS fare modelling and selected specification extensions. The original fares~v1 model (\texttt{fare\_attributes.txt} and \texttt{fare\_rules.txt}) is demonstrated using the Porto Andante zone-based tariff, with explicit identification of its limitations regarding passenger categories, fare media, and multi-operator scenarios. The GTFS Fares~v2 schema (\texttt{fare\_products.txt}, \texttt{fare\_media.txt}, \texttt{rider\_categories.txt}, \texttt{fare\_capping.txt}) is then presented as the structured solution to those limitations. The module also covers GTFS-Flex for demand-responsive services---relevant to Portuguese intermunicipal operators---and introduces GTFS-Realtime as the dynamic complement to the static feed.

    \item \textbf{Practical Tooling:} Equips readers with the production and analysis toolkit needed to operate a GTFS feed in a real-world context. It covers MobilityData Canonical GTFS Validator usage and interpretation of common Portuguese feed errors; Python-based analysis using \texttt{partridge} and \texttt{pandas} for temporal and spatial coverage assessment; visualisation platforms; a reference ETL pipeline architecture for automated feed generation; and recommendations for feed automation via scheduled tasks and continuous integration pipelines. The module closes with guidance on concurrent GTFS and NeTEx output from a single source pipeline, anticipating dual-format obligations under the Portuguese national access point requirements.
\end{enumerate}

Each module includes worked CSV examples derived from the STCP Line~200 dataset, practical exercises with annotated solutions, best-practice callouts aligned with the MobilityData GTFS Schedule Best Practices, and a formal references section. By grounding every concept in Portuguese operational data, embedding regulatory context throughout, and providing systematic NeTEx bridges, the documentation prepares transport practitioners to produce, validate, and publish conformant GTFS feeds while building the conceptual foundation required for the NeTEx transition.
